---
title: "Differential expression results"
author: "Andreas Gschwind"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Goal
Exploration of differential expression results for sample: `r snakemake@wildcards$sample`.

```{r requiredPackages, warning=FALSE, message=FALSE}
library(tidyverse)
library(cowplot)
```

```{r loadData}
# load files containing differential expression results
results <- read_tsv(snakemake@input[[1]], show_col_types = FALSE)

# add column indicating significant CRE-gene pairs
results <- mutate(results, significant = pval_adj < snakemake@params$fdr_threshold)
```

***

### TSS targeting controls
If TSS targeting perturbations were included as positive controls, these can be used to assess how
well the differential expression testing pipeline is well calibrated. If TSS control perturbation id
is the same as the target gene id, then this information is used to highlight intended targets. If
no TSS targeting controls are included, no plots are shown.

```{r, fig.height=4, fig.width=6}
# label for TSS controls
tss_ctrl_label <- snakemake@params$tss_ctrl_label

# plot TSS targeting controls if there are any
if (tss_ctrl_label %in% results$target_type) {
  
  # get TSS targeting controls and add a column specifying if the gene corresponds to the intended 
  # target (i.e is the gene of the perturbed TSS)
  tss_results <- results %>% 
    filter(target_type == tss_ctrl_label) %>% 
    mutate(intended_target = perturbation == gene)
  
  # get maximum absolute log-fold change
  max_lfc <- max(abs(tss_results$logFC), na.rm = TRUE)
  
  # create volcano plot for TSS controls
  ggplot(tss_results, aes(x = logFC, y = -log10(pval_adj), color = intended_target)) +
    geom_point(data = filter(tss_results, intended_target == FALSE), alpha = 1) + 
    geom_point(data = filter(tss_results, intended_target == TRUE), alpha = 1) + 
    geom_hline(yintercept = -log10(0.05), linetype = "dashed") + 
    geom_vline(xintercept = 0, linetype = "dashed") +
    scale_color_manual(values = c("TRUE" = "steelblue", "FALSE" = "gray42")) +
    scale_x_continuous(limits = c(-max_lfc, max_lfc)) +
    labs(title = "All TSS targeting controls", y = "-log10(FDR)", color = "Intended\ntarget") +
      theme_bw()

}
```

### Perturbation - gene pairs
Differential expression results for non-TSS targeting controls are visualized.

```{r, fig.height=3.5, fig.width=10, warning=FALSE}
# remove any TSS controls
results_no_TSSctrls <- filter(results, target_type != tss_ctrl_label)

# get maximum absolute log-fold change
max_lfc <- max(abs(results_no_TSSctrls$logFC), na.rm = TRUE)

# make volcano plot with all pairs
p1 <- ggplot(results_no_TSSctrls, aes(x = logFC, y = -log10(pval_adj), color = significant)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed") + 
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_color_manual(values = c("TRUE" = "firebrick3", "FALSE" = "gray42")) +
  scale_x_continuous(limits = c(-max_lfc, max_lfc)) +
  labs(title = "Signficance vs. effect size", y = "-log10(adj. p-value)") +
  theme_bw()

# plot -log10 p-value vs distance to TSS
p2 <- ggplot(results_no_TSSctrls, aes(x = abs(dist_to_tss) / 1000, y = -log10(pval_adj),
                              color = significant)) +
  geom_point(data = filter(results_no_TSSctrls, significant == FALSE), alpha = 0.5) +
  geom_point(data = filter(results_no_TSSctrls, significant == TRUE), alpha = 0.5) +
  labs(title = "Significance vs. distance to TSS", x = "Distance to TSS (kb)",
       y = "-log10(adj. p-value)") +
  scale_color_manual(values = c("TRUE" = "firebrick3", "FALSE" = "gray42")) +
  theme_bw()

plot_grid(p1, p2, nrow = 1)
```
